Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2012-02-16 boettiger.tex
adding knitted appendices

%Differing causes of the transition.
Even when we can exclude or ignore other dynamics and
restrict ourselves to systems that can produce a saddle-node bifurcation,
 approaches based on critical slowing down or rising variance (REFS) must further assume that a changing parameter
has brought the system closer to the bifurcation.
This assumption excludes at least three alternative explanations for the transition.
The first possibility is that a large perturbation of the system state
has moved the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.
This is an exogenous forcing that does not arise from the system dynamics, so it is not the kind of event we can expect to forecast.
(An example might be a sudden dramatic increase in fishing effort that pushes the population past a threshold.)
The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary~\citep{Ditlevsen2010}.
As discussed in \citet{Livina2012}, such noise induced transitions cannot be predicted through early warning signals.
The third scenario is that the system does pass through a saddle-node bifurcation,
but rather than gradually and monotonically approaching the critical point, the
bifurcation parameter moves in a rapid or highly non-linear way, making the detection of any gradual trend impossible.

@@

\section{A model based approach}
Model-based approaches are beginning to play a larger role in early warning signal detection,
though we have not as yet seen the direct fitting and simulation of models to compare hypotheses.
While choosing appropriate models without system-specific knowledge is challenging,
much can be accomplished by framing the implicit assumptions into equations.
\citet{Lade2011} introduces the idea of generalized models for early warning signals, and
\citet{Kuehn2011} presents normal forms for bifurcation processes that can give rise to critical transitions.
\citet{Carpenter2011e} and \citet{Dakos2011a} start by assuming the dynamics obey a generic
stochastic differential equation (SDE), but use this only to derive or define the summary statistics of interest.

@@

Asa way to clearly illustrate the choices involved in looking for warning signals while avoiding false alarms, we introduce an approach based on receiver operator curves. These curves XXX. The extent of the difficulty in finding reliable indicators of impending regime shifts based on summary statistics. becomes clear from the ROC curves of these statistics,
where a 5\% false positive rate often corresponds to only a 5\% true positive rate,
performing no better than the flip of a coin. %%%% EEEK are you really going to say that?!?!
By estimating the ROC curve for a given set of data,
we can better avoid applying warning signals in cases of inadequate power.
By taking advantage of the assumptions being made to write down a specific likelihood function,
we can develop approaches that get the most information from the data available.

DISUCSS EXAMPLES HERE

%%% Model adequacy
In any application of early warning signals, it is essential to address the question of model adequacy.
Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.
As the bifurcation results from the principle eigenvalue passing through zero,
the warning signal is expected in linear-order dynamics;
estimation of the nonlinear model is less powerful and less accurate.
The performance of this approach in the simulated data -- which is nonlinear in its dynamics
and driven with non-Gaussian noise introduced by the Poisson demographic events --
demonstrates the accuracy under violation of these assumptions.

>>>>

%Differing causes of the transition.
Even when we can exclude or ignore other dynamics and
restrict ourselves to systems that can produce a saddle-node bifurcation,
approaches based on critical slowing down or rising variance~\citep[\emph{e.g.}][]{Held2004, Scheffer2009, Carpenter2011}
must further assume that a changing parameter has brought the system closer to the bifurcation.
This assumption excludes at least three alternative explanations for the transition.
The first possibility is that a large perturbation of the system state
has moved the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.
This is an exogenous forcing that does not arise from the system dynamics, so it is not the kind of event we can expect to forecast.
(An example might be a sudden dramatic increase in fishing effort that pushes the population past a threshold.)
The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary~\citep{Ditlevsen2010}.
As discussed in \citet{Livina2012}, such noise induced transitions cannot be predicted through early warning signals.
The third scenario is that the system does pass through a saddle-node bifurcation,
but rather than gradually and monotonically approaching the critical point, the
bifurcation parameter moves in a rapid or highly non-linear way, making the detection of any gradual trend impossible.

@@

\section{A model based approach}
Model-based approaches are beginning to play a larger role in early warning signal detection,
though we have not as yet seen the direct fitting and simulation of models to compare hypotheses.
While choosing appropriate models without system-specific knowledge is challenging,
much can be accomplished by framing the implicit assumptions into equations.
\citet{Lade2012} introduces the idea of generalized models for early warning signals, and
\citet{Kuehn2011} presents normal forms for bifurcation processes that can give rise to critical transitions.
\citet{Carpenter2011e} and \citet{Dakos2011a} start by assuming the dynamics obey a generic
stochastic differential equation (SDE), but use this only to derive or define the summary statistics of interest.

@@

 % past tense or not?
We have used three examples to illustrate the performance of this approach in data from 
simulation, a chemostat experiment, and paleo-atmospheric record; examples differing in 
sampling intensity and strength of signal of an approaching collapse.  
While the well-sampled geological data shows an unmistakable signal in this model-based approach,
the uncertainty in the smaller simulated and experimental data forces a trade-off between errors.  
%Because such uncertainty may be more typical of early warning signal detection 
%than the monotonically increasing patterns often used to illustrate the approach,

As a way to clearly illustrate the choices involved in looking for warning signals while avoiding false alarms,
we introduce an approach based on receiver operator curves. 
These curves illustrate the extent to which an potential warning signal mitigates the trade-off 
between missed events and false alarms.  
The extent of the difficulty in finding reliable indicators of impending regime shifts
based on summary statistics becomes clear from the ROC curves of these statistics,
where a 5\% false positive rate often corresponds to only a 5\% true positive rate,
performing no better than the flip of a coin. %%%% EEEK can we really say it that strongly?
By estimating the ROC curve for a given set of data,
we can better avoid applying warning signals in cases of inadequate power.
By taking advantage of the assumptions being made to write down a specific likelihood function,
we can develop approaches that get the most information from the data available.

%%% Model adequacy
In any application of early warning signals, it is essential to address the question of model adequacy.
Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.
As the bifurcation results from the principle eigenvalue passing through zero,
the warning signal is expected in linear-order dynamics;
estimation of the nonlinear model is less powerful and less accurate.
The performance of this approach in the simulated data -- which is nonlinear in its dynamics
and driven with non-Gaussian noise introduced by the Poisson demographic events --
demonstrates the accuracy under violation of these assumptions.
% this robustness to assumptions can also be seen in the performance of the empirical systems
